Classical Planning Algorithms ¡ on the Atari Video Games: Preliminary Results

### 2014-10-31 workshop-paper/paper.tex
inputs removed

\newcommand{\Omit}[1]{}

\newcommand{\eqdef}{\stackrel{\hbox{\tiny{def}}}{=}}
\newcommand{\IW}{{\textit{IW}}}
\newcommand{\SIW}{{\textit{SIW}}}
\newcommand{\ID}{{\textit{ID}}}
\newcommand{\BRFS}{{\textit{BrFS}}}
\newcommand{\BFS}{{\textit{2BFS}}}

@@

\input{performance_table}

\input{experimental_results}

>>>>

\newcommand{\Omit}[1]{}
%\newcommand{\propername}[1]{\text{\textsf{\small #1}}\xspace}
\newcommand{\propername}[1]{\textsc{#1}}
\newcommand{\eqdef}{\stackrel{\hbox{\tiny{def}}}{=}}
\newcommand{\IW}{{\textit{IW}}}
\newcommand{\SIW}{{\textit{SIW}}}
\newcommand{\ID}{{\textit{ID}}}
\newcommand{\BRFS}{{\textit{BrFS}}}
\newcommand{\BFS}{{\textit{2BFS}}}
\newcommand{\UCT}{{\textit{UCT}}}
\newcommand{\Asterix}{\propername{Asterix}}
\newcommand{\BeamRider}{\propername{Beam Rider}}
\newcommand{\Breakout}{\propername{Breakout}}
\newcommand{\Enduro}{\propername{Enduro}}
\newcommand{\Freeway}{\propername{Freeway}}
\newcommand{\Pong}{\propername{Pong}}
\newcommand{\Qbert}{\propername{Q*Bert}}
\newcommand{\Seaquest}{\propername{Seaquest}}
\newcommand{\SpaceInvaders}{\propername{Space Invaders}}
\newcommand{\Deepmind}{\propername{Deepmind}}

@@

\begin{table*}
\addtolength{\tabcolsep}{-2pt}
{\scriptsize
\begin{tabular}{@{}l|rr|rrr|rrr|rrr|rrr|c@{}}
%% table header
\hline\\[-3ex]
\multicolumn{1}{c|}{} & \multicolumn{2}{c|}{Random} & \multicolumn{3}{c|}{\BRFS}& \multicolumn{3}{c|}{$\IW(1)$}& \multicolumn{3}{c|}{\BFS}& \multicolumn{3}{c|}{\UCT}
& \multicolumn{1}{c}{\Deepmind} \\[.2ex]
\hline\\[-3ex]
%% table body
\multicolumn{1}{c}{Game}  & \multicolumn{1}{c}{S} & \multicolumn{1}{c|}{SD}  & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{SD} &  \multicolumn{1}{c|}{T}  & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{SD} & \multicolumn{1}{c|}{T}  & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{SD} & \multicolumn{1}{c|}{T}  & \multicolumn{1}{c}{S} & \multicolumn{1}{c}{SD} & \multicolumn{1}{c|}{T} & \multicolumn{1}{c}{S}\\[.2ex]
\hline\\[-3ex]
\Asterix        & 210    & 54.77     & 3900  & 5538.95   & 32.24 & 119000        & 4987.48   & 17.08 & 56320       & 4757.84   & 0.95  & \textbf{296300 (290700)} & 6657.7    & 10.82 & n/a\\
\BeamRider     & 448.8  & 84.64     & 519.2 & 179.81    & 43.48 & 3000          & 1150.22   & 6.02  & 3804        & 446.18    & 0.92  & \textbf{4320 (6624)}     & 2175.26   & 42.91 & 4092\\
\Breakout       & 1      & 0.71      & 2     & 1.41      & 51.56 & 383.6         & 9.84      & 1.37  & 421.2       & 13.66     & 3.95  & \textbf{512.4 (364)}     & 193.97    & 26.03 & 168\\
\Enduro          & 0      & 0         & 0.4   & 0.89      & 40.92 & \textbf{500}  & 0         & 38.13 & 292.8       & 6.18      & 13.12 & 306.8 (286)              & 8.14      & 32.14 & 470\\
\Freeway         & 0      & 0         & 0     & 0         & 71.84 & \textbf{29.4} & 0.89      & 15.19 & 19.6        & 3.29      & 4.44  & 1 (0)                    & 1.41      & 90.64 & n/a \\
\Pong           & -20.4  & 0.55      & -20.2 & 0.84      & 41.51 & 16.4          & 1.67      & 18.45 & \textbf{21} & 0         & 39.69 & \textbf{21 (21)}         & 0         & 38.37 & 20 \\
\Qbert           & 185    & 345.33    & 160   & 85.88     & 42.83 & 1545          & 1666.06   & 1.82  & 14725       & 7872.1    & 1.76  & \textbf{18730 (17343)}   & 1127.69   & 28.61 & 1952\\
\Seaquest       & 192    & 38.99     & 200   & 14.14     & 40.01 & \textbf{4274} & 4206.81   & 14.72 & 4070        & 1939.55   & 1.59  & 3406 (5132)              & 2402.47   & 26.16 & 1705 \\
\SpaceInvaders & 187    & 120.86    & 115   & 61.44     & 46.11 & \textbf{2670} & 637.3     & 12.1  & 1327        & 598.04    & 2.25  & 2634 (2718)              & 375.66    & 19.66 & 581
\end{tabular}
}

 \caption{\small Random baseline, \BRFS, $\IW(1)$, \BFS, \UCT~performance over 5 runs,
   setting the Atari 2600 random seeds to $0, \ldots, 4$, and maximum episode duration of $18,000$
   frames. \textit{T} is the average CPU time per call to each action selection algorithm, \textit{S} is the average score achieved,
   and \textit{SD} is the standard deviation. \UCT~performance
   reported in \cite{bellemare:jair2013} is shown in parenthesis. Numbers
   in bold show best performer in terms of average score. Average score reported in 
   \cite{deep-mind-atari} is reported for reference under the column \Deepmind.
   }
\label{performance}

\end{table*}

We have tested the $\IW(1)$ and \BFS~described in the previous Section
over $9$ games selected amongst those considered in the \emph{training} set
of \cite{bellemare:jair2013}, and those discussed in \cite{deep-mind-atari}. 
In Table~\ref{performance} the performance of both $\IW(1)$ 
and \BFS, characterized by the average score obtained over all episodes, the standard deviation
of the score obtained and time (in seconds) per lookahead, is compared with that of \UCT, 
\BRFS~and the baseline random action selection. Our experimental 
setup  follows that of \cite{bellemare:jair2013}, with three exceptions. First, rather than
considering $10$ episodes of each game, we have considered just $5$
due to time constraints, as each game can take as much as $40$ hours of 
CPU time. Second, while \UCT~is limited to perform up to $500$ base policy rollouts ($150,000$ frames
are to be simulated while performing the lookahead), both $\IW(1)$ and \BFS are limited to
simulate $100,000$ frames. Other \UCT~parameters such as the exploration constant $C$ and
the planning horizon $m$ have not been modified ($C=0.1$, $m=300$). Thirdly and last, both $\IW(1)$ 
and \BFS~are limited to a planning horizon $m=1500$, where one only an action is evaluated every $5$ frames, matching
the setup used during actual gameplay (an action is selected every $5$ frames, using a no-op
for the remaining $4$). We include the average score reported for \UCT~by \cite{bellemare:jair2013} 
in Table~\ref{performance} bracketed with parenthesis for reference. Also unchanged is the discount factor
used by all the algorithms ($\gamma=0.995$). All algorithms reuse the tree built between each 
lookahead by setting the root of the tree to the best child of the root in the previous one,
and deleting its siblings and their descendants, as \cite{bellemare:jair2013} does for \UCT. We note that 
this may have a different effect on algorithms where all branches have the same number of
generated nodes, as is the case of \BRFS, than in $\IW(1)$, \BFS~and \UCT, where the depth of
the branches is uneven.

Table~\ref{performance} shows that both $\IW(1)$ and \BFS~outperform by a wide margin 
\BRFS, which sometimes performs worse than the baseline, both in terms of average score attained and 
average time per call. \BRFS~rarely collects any reward to inform the choice of action at
the root node of the lookahead, as the depth of the \BRFS~search tree, having limited the number of
frames allowed to be simulated to $100,000$, corresponds to a lookahead of $0.3$ seconds ($20$ 
frames or $4$ nodes). Both $\IW(1)$ and \BFS~reach states up to $350$--$1500$ frames ahead 
($70$--$260$ nodes or $6$--$22$ seconds), and that results in collecting enough rewards so as to
inform the choice at the root in a way that much higher scores are attained. Interestingly, very often
neither $\IW(1)$ or \BFS~fully consume the number of simulation frames allocated, since many
states are pruned, in the case of $\IW(1)$, or relegated to the back of the open lists, in the 
case of \BFS, due to having a novelty of $0$ and the planning horizon being limited to $m$.

We observe that in $4$ out of $9$ games, $\IW(1)$ outperforms \UCT~limited to the same number
of simulated frames, both in terms of average score attained and average time per call. \BFS~is
generally faster than $\IW(1)$, but only outperforms \UCT~in $2$ out of $9$ games (\Freeway and
\Seaquest) and $\IW(1)$ in $4$ out of $9$ games. Compared with the \UCT~results reported
in \cite{bellemare:jair2013}, that by limiting the number of rollouts allow a $50$\% more frames
to be simulated in each lookahead, $\IW(1)$ still outperforms \UCT~in $3$ out of $9$ games
(\Breakout, \Enduro~and \Freeway). As introduced in the previous Section, the order in which
$\IW(1)$ and \BFS~consider actions is randomized, since the novelty of a given state may
change depending on the order in which states are considered. We found this randomization to
have a significant effect in \Qbert~where not randomizing the order in which actions are considered
resulted in $\IW(1)$ being trapped in a loop. Last, we note that both $\IW(1)$ and \BFS~attain similar 
performance on \Freeway~as the baseline semi-random algorithm \emph{Perturb} 
from \cite{bellemare:jair2013}, which beated both \UCT~and the reinforcement learning algorithms reported.

We finish the discussion of the results presented in Table~\ref{performance} by observing
that $\IW(1)$ outperforms the average scores reported for the best algorithm reported in
\cite{deep-mind-atari} but in $4$ out of $7$ games, and \BFS~does so too in $5$ out of $7$
games.

We do not report $IW(2)$ in Table~\ref{performance} since we found its performance to
be very similar to that of \BRFS~in some preliminary tests. When considering $2$--tuples
of state variables to evaluate the novelty of states, it tended to result in a lookahead
$5$ times deeper than those resulting from applying \BRFS~, but still far too shallow so
as to compete with the other planning algorithms considered.
